\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\

### 2020-06-01 manuscript/manuscript.tex
manuscript

% Title
\title{
    \usefont{OT1}{bch}{b}{n}
    \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt]
    \huge Deep learning transcriptomic model for prediction of pan-cancer chemotherapeutic sensitivity \\
}
\selectlanguage{english}
\usepackage{authblk}
\author[1]{Eddie Guo}
\author[2]{Mehul Gupta}
\author[1]{Pouria Torabi}
\author[2]{Sunand Kannappan}
\affil[1]{University of Alberta}
\affil[2]{University of Calgary}

@@

@@

% Abstract
\begin{abstract}
	Emerging precision oncology studies have yet to generate a predictive biomarker that utilizes gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. This development would allow for identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. As such, this study utilized cell line sensitivity and molecular profiling data to generate a combinatorial gene expression predictive biomarker, utilizing feature selection and a deep learning model. A pan-cancer cohort of cell-line gene expression data from Genomics of Drug Sensitivity in Cancer (GDSC) was clustered into two response groups. Cell line response groups showed a significant difference in pan-drug chemotherapeutic sensitivity. Due to the high dimensional nature of the microarray data, biologically agnostic feature selection was conducted to identify genes with the highest predictive value. The feature space was reduced to 300 genes, which functional profiling indicated enriched primarily the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using these selected genes, a deep learning neural network architecture was developed to predict response groups. Hyperparameter tuning of the deep learning model dictated a 4 hidden layer architecture with a predictive accuracy of 91.7\%. \vspace{1em}

@@

\begin{multicols}{2}
% Introduction
\section{Introduction}
With the advent of high-throughput technology, precision oncology approaches have utilized molecular characteristics of tumours to inform clinical decision-making, including chemotherapeutic regimen. The major focus of these approaches has been the development of targeted therapeutics, which are selective for specific genetic aberrations and expression profiles. Although these approaches may be extremely successful for inducing tumour response, tumours are more likely to gain resistance to therapies with specific targets \cite{small_mol}. Furthermore, not all tumours present with targetable features \cite{small_mol}.

Emerging precision oncology approaches have begun to utilize high-throughput technology to potentiate the usage of conventional and less-targeted chemotherapy. Given that many of these less-targeted and consequently more cytotoxic chemotherapies have broad activity, the primary determinants of chemotherapeutic selection include cancer type and certain molecular markers \cite{adverse}. Nevertheless, it is well established that tumour sensitivity to chemotherapy is heterogeneous between, but also within cancer types – resulting in a subset of patients that fail to respond to cancer type-specific chemotherapy while being exposed to significant side effect burden \cite{adjuvant}. Given that evidence suggests that gene expression can mediate drug response, recent advances have utilized individual and combinatorial gene expression biomarkers to develop predictors of tumour sensitivity to chemotherapeutic compounds \cite{integrated}.

While previous studies have developed predictive biomarkers for specific drugs, the utility of these biomarkers is limited to the determination of whether specific drugs may be effective for a patient \cite{drug_sense}. However, a pan-drug predictive biomarker may provide significant clinical utility in the selection of candidate drugs for a certain patient for further investigation. Such a biomarker could be developed if pan-cancer tumours with similar expression have similar drug responses across all chemotherapies, with few exceptions. The availability of pan-cancer cell line databases with in-vitro drug sensitivity analyses along with accompanying gene expression profiling provides an ideal model for such analyses \cite{gdsc}. However, most previous drug sensitivity predictive biomarkers built on cell line databases have utilized classical machine learning combinatorial techniques - which fail to capture the dimensionality of available transcriptomic data. However, advanced deep learning algorithm approaches are capable of handling such dimensionality fail to allow interpretability, and consequently include transcriptomic data volume that is clinically infeasible \cite{ml_oncol}. As such, deep learning algorithms that utilize feature selection processes may maximize both the accuracy and functionality of such predictive biomarkers.

As such, we set out to generate a combinatorial gene expression predictor of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types – to i) demonstrate that gene expression influences chemotherapeutic response across most drugs, and to ii) motivate future studies into the development of a clinically applicable predictor of candidate chemotherapeutics for tumours of a specific gene expression profile.

Following unsupervised clustering of cell lines into therapeutic response groups with similar pan-drug sensitivity, we show that cancer type does not stratify cell lines by therapeutic response. We utilize a feature selection algorithm to iteratively select and identify a subset of 300 relevant genes that influence chemotherapeutic sensitivity. We then generate a combinatorial predictive model from the feature selected genes utilizing deep neural networks.

@@

Here we develop a deep learning model to accurately classify cancer cell lines  into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) consortium. Following data collection and curation, we utilized unsupervised learning algorithms to define two groups based on chemotherapeutic response. Next, we used a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. We created an optimized neural network that utilizes transcriptomics features to classify patients into therapeutic response groups. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline.

\subsection*{Pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell-line and associated therapeutic efficacy data were obtained from the Genomics of Drug Sensitivity in Cancer (GDSC) database. This database includes 1,110 cell lines from various different tumour types, and is thought to represent a relatively comprehensive pan-cancer dataset. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cells cultured, with lower values being associated with improved drug efficacy. The data was used to generate a matrix with cell-line and accompanying therapeutic information. This dataset was filtered to exclude therapies with less than 80\% of data for all cell-lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step. This resulted in the inclusion of 548 cell lines (49.4\% of the original cell lines) and 117 (46.6\%) therapeutics for clustering based analysis.

% flowchart figure
\begin{figure*}[!ht]
    \centering
    \begin{tikzpicture}[node distance=6em, auto]
        % blocks
        \node [block, text width=24em] (init) {Collect data from GDSC (1,110 cell lines, 251 drugs).};
        \node [block, below of=init, text width=24em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC50 via z-score.};
        \node [block, below of=process, text width=24em, yshift=-1.25em] (kmeans) {K-means clustering into 2 response groups (verified via elbow graph).};
        \node [block, below of=kmeans, text width=24em, yshift=-0.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC50 matrix.};
        \node [block, below of=append, text width=24em, yshift=-1em] (feat) {Feature selection for relevant genes in RNA-seq data via Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
        \node [block, below of=feat, text width=24em, yshift=-2.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter optimization via grid search. Models evaluated via 5-fold cross-validation.};
        \node [block, below of=neural, text width=24em, yshift=-1.5em] (annot) {Gene ontology (GO) annotation of predictive genes.};

@@

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic response matrices obtained from the GDSC consortium were used to evaluate conventional tools used to separate patients into chemotherapeutic response groups as well as bifurcate candidate cell lines into defined response cohorts. Classical cancer classifications, including anatomic location and solid vs. non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status were used to evaluate whether classical classification systems were able to stratify cell lines into chemotherapeutic response groups \cite{colorectal}, \cite{gi}, \cite{lung}, \cite{breast}. Cell-lines were separated into subgroups based on anatomic location, solid vs. non-solid tumour status, and mutation/non-mutation groups for TP53 and KRAS respectively. These subgroups were then plotted with the first and second principal components to identify if they clustered together. Observed clustering would be indicative of these classifiers being predictive of chemotherapeutic response.

Following evaluation of existing classifiers, we attempted to create defined cell-line clusters on the basis of the observed chemotherapeutic response of the pan cancer cell-line sample. We developed a Euclidean distance matrix for the retained cell lines based upon their pan-chemotherapy response. This matrix was then used to identify an optimal number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines. K-means clustering was then utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the Pheatmap package in R. Separation between clusters was also visualized using principal component analysis with the factoextra package in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare the half-maximal inhibitory concentration (IC$_{50}$) values between the groups. We attempted to select the least number of clusters that retained significant differences in IC$_{50}$ values. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

% clustering figure
\begin{figure*}[!ht]
    \centering
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[height=2.75in]{Figures/pca.png}
        \caption{}
        \label{fig:pca}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[height=2.75in]{Figures/volcano.png}
        \caption{}
        \label{fig:volcano}
    \end{subfigure}

@@

    \caption{(\ref{fig:pca}) Principal component component analysis of pan-cancer cell line therapeutic efficacy, generated from IC$_{50}$ values of all available chemotherapeutics. The horizontal axis shows the first principal component, the vertical axis the second component. The two identified therapeutic response clusters are indicated in red and blue respectively. (\ref{fig:volcano}) Volcano plot identifying chemotherapeutics with significantly different IC50 values between therapeutic response clusters. Drugs identified in red meet the criteria for significance (FDR adjusted $p<0.05$). (\ref{fig:heatmap}) Heatmap of therapeutic IC50 for the two identified therapeutic response clusters. Columns represent individual chemotherapies, and are clustered according to euclidean distance. Colours range from yellow to purple, with a shift toward the latter indicating increased efficacy of the corresponding chemotherapeutic.}
    \label{fig:clustering}
\end{figure*}

\subsection*{Feature Selection}
To develop a transcriptomic model predictive of therapeutic response clusters, expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line was obtained. Minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610), and processed using the affy package in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster 1 and 5 from cluster 2), resulting in the inclusion of 541 cell-lines in model generation. The microarray dataset screened the expression levels of 16,382 genes, a model based on that many features is highly likely to overfit, compromising the generalizability of the model on new data. Such a large feature space also adds unnecessary noise and severely limits the accuracy and computational efficiency of the model \cite{liu}. Our approach to address these issues was dimensionality reduction through feature selection. The BorutaPy package in Python 3 is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}.

@@

    \caption{(\ref{fig:cms}) From left to right: confusion matrices for the 3, 4, and 5 hidden layer neural network models evaluating the true positive, false positive, true negative, and false negative rate. The models classify patient RNA-seq datasets into chemotherapy response cohorts. (\ref{fig:roc}) ROC curves for 3, 4, and 5 hidden layers neural network models with confidence bands of $\pm 1$ standard deviation. The models classify patient RNA-seq datasets into chemotherapy response cohorts. Each model was subject to 5-fold cross-validation and the mean across all trials was plotted.}
\end{figure*}

\subsection*{Classification using an optimized neural network}
The neural net was constructed using the Tensorflow Keras sequential deep learning API in Python 3. The model underwent multiple instances of optimization, starting with the manipulation of the overall hidden layer architecture. The classifier’s predictive accuracy and misclassification rate were monitored to determine the optimal number of dense hidden layers (Fig. \ref{fig:cms}) in addition to iterative manipulation of the number of neurons in each hidden layer. The rectified linear unit (ReLu) was chosen as the neuronal activation function for all the layers except the output layer which used a sigmoid activation as a means of classifying instances into binary classes.

The model was rigorously monitored for and protected against overfitting on the training dataset. To address this, we employed dropout layers with a 0.3 dropout rate and batch normalization layers were employed to improve the generalizability of the model.

The dataset was randomly segregated using the Pareto principle where we reserved 80\% of the data for training and the remaining 20\% for validation \cite{pareto}. Model selection was performed by tuning the model’s hyperparameters via a grid search and 5-fold cross-validation (Table \ref{tab:params}). We performed a grid search with 3-fold cross-validation on the training data (80\% of the dataset; 432 training samples, 541 overall) to determine the parameters under which to minimize the binary cross-entropy loss function. GridSearchCV from the scikit-learn library was used as a means of iterating through multiple possibilities of epochs, batch size, optimizer, and kernel initializer to find the optimal model. To prevent class imbalance during training, we used the Synthetic Minority Oversampling Technique (SMOTE) from the imblearn package for Python 3. Each model’s performance was evaluated by the area under the receiver operating characteristic (ROC) curve (AUC). Performance evaluation of the final model was performed with the testing set.

@@

\subsection*{Clustering of pan-cancer cell lines identifies two distinct therapeutic response cohorts}
We assess the ability of common molecular and clinical characteristics used in clinical decisions to stratify cell lines into groups with similar chemotherapeutic performance by subgrouping cell lines based upon these criteria and plotting them against the first and second principal components. Commonly used measures including the system of origin, molecular subtype, as well as TP53, and KRAS mutation status failed to identify defined clusters of cells with similar therapeutic responses (Fig. \ref{fig:overall_pca}). This likely indicates that these widely used classifiers may be inadequate to  stratify patients, and thus improved therapeutic classification system may prove useful for clinical and research settings.

To identify defined cohorts of pan-cancer cell-lines with similar trends in therapeutic efficacy, we employed unsupervised clustering of retained cell-lines. Principal component analysis was used to reduce the dimensionality of the dataset, allowing for visualization of defined therapeutic response cohorts. This process identified two distinct clusters of therapeutic efficacy (Fig. \ref{fig:pca}), 362 cell lines identified in response group 1, and 186 cell lines identified in response group 2.  Further analysis of these clusters demonstrate that a subset of therapeutics performs substantially differently between the different cohorts (Figure x - heatmap). To quantify differences in therapeutic response between clusters, IC50 values were compared between candidate cell lines (Fig. \ref{fig:volcano}). Of the 117 therapies included, 95 had significant differences in efficacy between the two cohorts identified. This suggests that these cohorts represent groups of cell lines with vastly different therapeutic responses. Therefore the ability to accurately stratify into these cohorts may be a valuable tool for stratification prior to chemotherapeutic treatment.

\subsection*{Boruta selects 300 genes}
The Boruta feature selection algorithm identified 300 relevant genes from the original set of 16,382 genes at $\alpha=0.05$ with a maximum tree depth of 5. Of importance, the algorithm selects genes that are estimated to have highest predictive value rather than biological significance. This biologically agnostic feature selection was key to limiting preliminary bias in the model. Functional profiling indicated that the top three pathways enriched by the selected genes were pathways in focal adhesion, extracellular matrix-receptor, and proteoglycan interactions in cancer.

% gene annotation figure
\begin{figure*}[!ht]
	\centering
	\includegraphics[width=\textwidth]{Figures/kegg.png}
	\caption{KEGG pathway functional enrichment for predictive genes included in the deep learning model conducted using g:Profiler.}
	\label{fig:kegg}
\end{figure*}

\subsection*{A neural network with four hiddeen layers accurately classifies patients into responder and non-responder cohorts}
Unsupervised learning in the form of the K-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Using these distinct therapy response cohorts, we developed a deep learning binary classification to predict drug response based on transcriptome data. We initially analyzed five neural network architectures, each corresponding to 1-5 hidden layers (Fig. \ref{fig:neural_vis}). Hyperparameter optimization via grid search returned similar results for each model: 50 epochs, batch size of 32, Adagrad as the optimizer, and a normal kernel initializer. Neural network architectures containing 3-5 hidden layers performed similarly with approximately 90\% accuracy. The architectures with 1 and 2 hidden layers performed less optimally with ~80\% accuracy. We proceeded to validate the architectures with 3-5 hidden layers using 5-fold cross-validation. Of note, the model with 4 hidden layers had the lowest false positive rate (FPR = 3.67\%) and a false negative rate (FNR = 4.59\%). The model with five hidden layers had the highest FPR of the models evaluated (8.26\%).

A receiver operating characteristic (ROC) curve was plotted for each of the neural network variants as an alternative evaluative method under uneven class sizes (Fig. \ref{fig:roc}). The relative ratio between the model’s false positive classification rate and its true positive rate was averaged between 5 K-Folds. The mean Area Under Curve (AUC) for the 5 trials was used to compare the three network architectures. However, the differences between the various architectures for AUC was not significatly different. Consequently, confusion matrices (Fig. \ref{fig:cms}) and associated misclassification rates were used to pick the optimal model. The neural net with 4 hidden layers boasted the best performance overall with a 91.7\% accuracy and an 8.3\% misclassification rate.

>>>>

% Title
\title{
    \usefont{OT1}{bch}{b}{n}
    \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt]
    \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\
}
\selectlanguage{english}
\usepackage{authblk}
\author[1]{Eddie Guo}
\author[2]{Mehul Gupta}
\author[1]{Pouria Torabi}
\author[2]{Sunand Kannappan}
\affil[1]{University of Alberta}
\affil[2]{University of Calgary}

@@

\setlength{\parskip}{0pt}

@@

% Abstract
\begin{abstract}
	Emerging precision oncology studies have yet to generate a predictive biomarker that utilizes gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. This development would allow for identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. As such, this study utilized cell line sensitivity and molecular profiling data to generate a combinatorial gene expression predictive biomarker, utilizing feature selection and a deep learning model. A pan-cancer cohort of cell-line gene expression data from Genomics of Drug Sensitivity in Cancer (GDSC) was clustered into two response groups. Cell line response groups showed a significant difference in pan-drug chemotherapeutic sensitivity. Due to the high dimensional nature of the microarray data, biologically agnostic feature selection was conducted to identify genes with the highest predictive value. The feature space was reduced to 300 genes, which functional profiling indicated was primarily enriched for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using these selected genes, a deep learning neural network architecture was developed to predict response groups. Hyperparameter tuning of the deep learning model dictated a 4 hidden layer architecture with a predictive accuracy of 91.7\%.  This validates the postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it suggests the potential utility of similar combinatorial biomarkers for selection of potent candidate drugs. \vspace{1em}

@@

\begin{multicols*}{2}
% Introduction
\section{Introduction}
With the advent of high-throughput sequencing technology, precision oncology approaches have utilized molecular characteristics of tumours to inform clinical decision-making, including choice of chemotherapeutic regimen. The major focus of these approaches has been the development of targeted therapeutics, which are selective for specific genetic aberrations and expression profiles. Although these approaches may be successful for inducing tumour response, tumours are more likely to gain resistance to therapies with specific targets \cite{small_mol}. Moreover, not all tumours present with targetable features \cite{small_mol}.

Emerging precision oncology approaches have begun to utilize high-throughput technology to potentiate the usage of conventional and less-targeted chemotherapy. Given that many of these less-targeted and consequently more cytotoxic chemotherapies have broad activity, the primary determinants of chemotherapeutic selection include cancer type and certain molecular markers \cite{adverse}. Nevertheless, it is well established that tumour sensitivity to chemotherapy is heterogeneous both between and within cancer types, which results in a subset of patients that fail to respond to conventional chemotherapy regimens while being subjected to significant side effect burden \cite{adjuvant}. Given that evidence suggests that gene expression can mediate drug response, recent advances have utilized individual and combinatorial gene expression biomarkers to develop predictors of tumour sensitivity to chemotherapeutic compounds \cite{integrated}.

While previous studies have developed predictive biomarkers for specific drugs, the utility of these biomarkers is limited to particular patients and clinical contexts \cite{drug_sense}. That is, these studies are limited in terms of clinical generalizability to different chemotherapeutic regimens. However, a pan-drug predictive biomarker may provide significant clinical utility in the selection of candidate therapies for particular patients. Such a biomarker could be developed if pan-cancer tumours with similar expression have similar drug responses across all chemotherapies, with few exceptions. The availability of pan-cancer cell line databases with \textit{in vitro} drug sensitivity analyses along with accompanying gene expression profiling provides an ideal model for such analyses \cite{gdsc}. However, most previous drug sensitivity predictive biomarkers built on cell line databases have utilized classical machine learning combinatorial techniques, which fail to capture the dimensionality of available transcriptomic data. However, advanced deep learning algorithmic approaches that are capable of handling such dimensionality often fail to allow interpretability, and consequently require transcriptomic data that is clinically infeasible \cite{ml_oncol}. Thus, deep learning approaches should minimize the number of transcriptomic features selected to maximize both the accuracy and functionality of such predictive biomarkers.

As such, we set out to generate a deep learning transcriptomic model for the prediction of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types. If successful, this would demonstrate that gene expression influences chemotherapeutic response across most drugs and further motivate future studies into the development of clinically applicable predictors of candidate chemotherapeutics for tumours of a specific gene expression profile.

Following unsupervised clustering of cell lines into therapeutic response groups with similar pan-drug sensitivity, we show that conventional clinical criteria fails to stratify cell lines by therapeutic response. We utilize a biologically agnostic feature selection algorithm to iteratively select and identify a subset of 300 relevant genes predictive of chemotherapeutic sensitivity. We then generate a combinatorial model from the selected genes utilizing neural networks that showcases a strong predictive ability across pan-cancer cell lines.

@@

Here we developed a deep learning model to accurately classify cancer cell lines into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) consortium. Following data collection and curation, we utilized unsupervised clustering algorithms to define two groups based on chemotherapeutic response. Next, we employed a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. We developed an optimized neural network that utilizes transcriptomics features to classify patients into therapeutic response groups. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline.

\subsection*{Pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell-line and associated therapeutic sensitivity data were obtained from the GDSC database. This database includes 1,110 cell lines from various tumour types, and is thought to represent a relatively comprehensive pan-cancer set. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cultured cells, with lower values being associated with improved drug efficacy. The data was used to generate a matrix with cell line and accompanying therapeutic information. This dataset was filtered to exclude therapies with less than 80\% of data for all cell lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step.

% flowchart figure
\begin{figure*}[!ht]
    \centering
    \begin{tikzpicture}[node distance=6em, auto]
        % blocks
        \node [block, text width=24em] (init) {Obtain data from the GDSC consortium (1,110 cell lines, 251 drugs).};
        \node [block, below of=init, text width=24em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC50 via z-score.};
        \node [block, below of=process, text width=24em, yshift=-1.25em] (kmeans) {Unsupervised clustering into 2 chemotherapeutic response groups.};
        \node [block, below of=kmeans, text width=24em, yshift=-0.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC50 matrix.};
        \node [block, below of=append, text width=24em, yshift=-1em] (feat) {Feature selection for relevant genes in RNA microarray data via Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
        \node [block, below of=feat, text width=24em, yshift=-2.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter optimization via grid search. Models evaluated via 5-fold cross-validation.};
        \node [block, below of=neural, text width=24em, yshift=-1.5em] (annot) {KEGG pathway enrichment of predictive genes.};

@@

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic sensitivity matrices were used to evaluate whether conventional clinical criteria could separate patients into previously-defined chemotherapeutic response groups. These conventional clinical criteria included anatomic location and solid versus non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status \cite{colorectal}, \cite{gi}, \cite{lung}, \cite{breast}. Cell-lines were separated into subgroups based on these criteria and plotted to determine whether these criteria effectively clustered response groups.

Following evaluation of existing classifiers, we attempted to create defined cell-line clusters on the basis of the observed chemotherapeutic response of the pan-cancer cell line sample. We developed a Euclidean distance matrix for the retained cell lines based upon their pan-chemotherapy response. This matrix was then used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. K-means clustering was then utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the Pheatmap package in R. Separation between clusters was also visualized using principal component analysis with the factoextra package in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

% clustering figure
\begin{figure*}[!ht]
    \centering
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[height=2.5in]{Figures/pca.png}
        \caption{}
        \label{fig:pca}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[height=2.5in]{Figures/volcano.png}
        \caption{}
        \label{fig:volcano}
    \end{subfigure}

@@

    \caption{(\ref{fig:pca}) Principal component analysis of pan-cancer cell line therapeutic efficacy, generated from IC$_{50}$ values of all available chemotherapeutics. The horizontal axis shows the first principal component, the vertical axis the second component. The two identified therapeutic response clusters are indicated in red and blue respectively. (\ref{fig:volcano}) Volcano plot identifying chemotherapeutics with significantly different IC$_{50}$ values between therapeutic response clusters. Drugs identified in red meet the criteria for significance (FDR adjusted $p<0.05$). (\ref{fig:heatmap}) Heatmap of therapeutic IC$_{50}$ for the two identified therapeutic response clusters. Columns represent individual chemotherapies and are clustered according to Euclidean distance. Colours range from yellow to black, with a shift toward the latter indicating increased efficacy of the corresponding chemotherapeutic.}
    \label{fig:clustering}
\end{figure*}

\subsection*{Feature Selection}
To develop a transcriptomic model predictive of therapeutic response clusters, expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line was obtained. Minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell-lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model based on that many features is highly likely to overfit, compromising the generalizability of the model on new data. Such a large feature space also adds unnecessary noise and severely limits the accuracy and computational efficiency of the model which was addressed using dimensionality reduction through feature selection via the BorutaPy package in Python 3  \cite{liu}. The BorutaPy package is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}. Computational efficiency and the resulting feature set quality were motivators for choosing Boruta over other selection algorithms such as univariate selection or principle component analysis.

@@

    \caption{(\ref{fig:cms}) From left to right: confusion matrices for the 3, 4, and 5 hidden layer neural network models evaluating the true positive, false positive, true negative, and false negative rate. The models classify cell line microarray datasets into chemotherapy response cohorts. (\ref{fig:roc}) ROC curves for 3, 4, and 5 hidden layers neural network models with confidence bands of $\pm 1$ standard deviations. Each model was subject to 5-fold cross-validation, and the mean ROC curve across all trials was plotted.}
\end{figure*}

\subsection*{Classification using an optimized neural network}
The neural net was constructed using the Tensorflow Keras sequential deep learning API in Python 3. The model underwent multiple instances of optimization, starting with the manipulation of the overall hidden layer architecture. The classifier's predictive accuracy and misclassification rate were monitored to determine the optimal number of dense hidden layers (Fig. \ref{fig:cms}) in addition to iterative manipulation of the number of neurons in each hidden layer. The rectified linear unit (ReLu) was chosen as the neuronal activation function for all the layers except for the output layer which used a sigmoid activation as a means of classifying instances into binary classes.

The model was rigorously monitored for overfitting on the training dataset. To minimize overfitting, we employed batch normalization layers followed by dropout layers with a 0.3 dropout rate to improve the generalizability of the model.

The dataset was randomly segregated using the Pareto principle where we reserved 80\% of the data for training and the remaining 20\% for validation \cite{pareto}. Model selection was performed by hyperparameter tuning using a grid search followed by 5-fold cross-validation (Table \ref{tab:params}). We performed a grid search with 3-fold cross-validation on the training data (80\% of the dataset; 432 training samples, 541 overall) to determine the parameters which minimize the binary cross-entropy loss function. GridSearchCV from the scikit-learn library was used as a means of iterating through multiple possibilities of epochs, batch size, optimizer, and kernel initializer to find the optimal model (Table \ref{tab:params}). To prevent class imbalance during training, we used the Synthetic Minority Oversampling Technique (SMOTE) from the imblearn package for Python 3. Each model’s performance was evaluated by the Area Under Curve (AUC) of the receiver operating characteristic (ROC) curve. Performance evaluation of the final model was performed with the testing set.

@@

\subsection*{Clustering of pan-cancer cell lines identifies two distinct therapeutic response cohorts}
From the GDSC consortium, we included 548 cell lines (49.4\% of the original cell lines) and 117 (46.6\% of the original drugs) therapeutics for response group clustering. We assessed the ability of common molecular and clinical characteristics to stratify cell lines into groups with similar chemotherapeutic performance by subgrouping cell lines based upon these criteria and plotting them against the first and second principal components. Commonly used measures including the anatomical location and morphologic subtype, as well as TP53 and KRAS mutation status failed to identify defined clusters of cells with similar therapeutic responses (Fig. \ref{fig:overall_pca}).

To identify defined cohorts of pan-cancer cell lines with similar trends in therapeutic sensitivity, we employed unsupervised clustering of retained cell-lines. Principal component analysis was used to reduce the dimensionality of the dataset, allowing for visualization of defined therapeutic response groups. This process identified two distinct clusters of therapeutic sensitivity (Fig. \ref{fig:pca}), 362 cell lines identified in response group A, and 186 cell lines identified in response group B. The cohorts perform substantially differently in a subset of therapeutics (Fig. \ref{fig:heatmap}). To quantify differences in therapeutic response between clusters, IC50 values were compared between candidate cell lines (Fig. \ref{fig:volcano}). Of the 117 therapies included, 95 had significant differences in efficacy between the two cohorts identified. This suggests that these cohorts represent groups of cell lines with vastly different therapeutic responses. Therefore the ability to accurately stratify into these cohorts may be a valuable tool for stratification prior to chemotherapeutic treatment.

\subsection*{Boruta selects 300 genes}
In order to select genes that are estimated to have the highest predictive value rather than biological significance, the Boruta feature selection algorithm was used. The feature selection algorithm identified 300 relevant genes from the original set of 16,382 genes at $\alpha=0.05$ with a maximum tree depth of 5. To better understand the transcriptomic heterogeneity underlying the therapeutic response cohorts, a KEGG pathway enrichment analysis was performed on feature-selected genes used in the deep learning model. This analysis identified enrichment of gene sets associated with focal adhesion and PI3K signalling among others (Fig. \ref{fig:kegg}).

% gene annotation figure
\begin{figure*}[!ht]
	\centering
	\includegraphics[width=\textwidth]{Figures/kegg.png}
	\caption{KEGG pathway functional enrichment for feature-selected genes included in the deep learning model. The vertical dotted line indicates the threshold for significance (adjusted $p < 0.05$).}
	\label{fig:kegg}
\end{figure*}

\subsection*{A neural network with four hidden layers accurately classifies patients into responder and non-responder cohorts}
Unsupervised learning in the form of the K-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Using these distinct therapy response cohorts, we developed a deep learning binary classifier to predict drug response groups based on transcriptome data. We initially analyzed five neural network architectures, each corresponding to 1-5 hidden layers (Fig. \ref{fig:neural_vis}). Hyperparameter optimization via grid search returned similar results for each model: 50 epochs, batch size of 32, Adagrad as the optimizer, and a normal kernel initializer. Neural network architectures containing 3-5 hidden layers performed similarly with approximately 90\% accuracy. The architectures with 1 and 2 hidden layers performed less optimally with approximately 80\% accuracy. We proceeded to validate the architectures with 3-5 hidden layers using 5-fold cross-validation. Of note, the model with 4 hidden layers had the lowest false positive rate (FPR, 3.67\%) and false negative rate (FNR, 4.59\%). The model with five hidden layers had the highest FPR (8.26\%) of the models evaluated.

An ROC curve was plotted for each of the neural network variants as an alternative evaluative method under uneven class sizes (Fig. \ref{fig:roc}). The relative ratio between the model’s FPR and its true positive rate (TPR) were averaged between 5 k-folds. The mean AUC for the 5 trials was used to compare the three network architectures. However, the differences between the various architectures for AUC was not significantly different. Consequently, confusion matrices (Fig. \ref{fig:cms}) and associated misclassification rates were used to pick the optimal model. The neural net with 4 hidden layers demonstrated the best performance overall with a 91.7\% accuracy.
